---
title: "Stats209Project"
output: html_document
date: "2023-12-08"
---

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(AER)
library(progress)
library(DOS2)
library(optmatch)
library(rcbalance)
library(RItools)
```


## Importing data
```{r}
pre <- read.csv("/Users/nataliehilderbrand/Downloads/sb_ca2019_all_csv_v4/sb_ca2019_all_csv_v4.txt")
post <- read.table("/Users/nataliehilderbrand/Downloads/sb_ca2021_all_csv_v2/sb_ca2021_all_csv_v2.txt", header = TRUE, sep = "^")
pre <- pre %>% filter(County.Code != 0 & District.Code != 0 & School.Code != 0 & Grade != 13)
post <- post %>% filter(County.Code != 0 & District.Code != 0 & School.Code != 0 & Grade != 13)
```

## Cleaning
```{r}
post <- subset(post, select = -Type.ID)
names(post)[names(post) == "Percentage.Standard.Met.and.Above"] <- "passed"
colnames(pre) <- colnames(post)

# drop columns we don't need
pre <- pre[, c("Student.Group.ID","Students.with.Scores", "Students.Enrolled", "Test.ID", "Grade", "Test.Year", "School.Code", "passed")]
post <- post[, c("Student.Group.ID","Students.with.Scores", "Students.Enrolled", "Test.ID", "Grade", "Test.Year", "School.Code", "passed")]
```





# All student groups: difference between 2019 and 2021 scores
## All student groups
```{r}
pre_all <- pre %>% filter(Student.Group.ID == 1)
post_all <- post %>% filter(Student.Group.ID == 1)

# drop null columns
pre_all <- pre_all[!is.na(pre_all$passed) & pre_all$passed!="" & pre_all$passed!="*",]
post_all <- post_all[!is.na(post_all$passed) & post_all$passed!="" & post_all$passed!="*",]

# add ID for matching with school code, grade, test type
pre_all$ID <- paste(pre_all$School.Code,pre_all$Grade,pre_all$Test.ID,sep="")
post_all$ID <- paste(post_all$School.Code,post_all$Grade,post_all$Test.ID,sep="")
```

## Matching on all student groups
```{r}
all <- data.frame(pre_all)
matches <- list()
match_scores <- list()
for (i in 1:(nrow(all))) {
  curr_id <- all[i,"ID"]
  new_val <- ifelse(any(post_all$ID == curr_id), post_all[post_all$ID == curr_id, "passed"], "*")
  scores <- ifelse(any(post_all$ID == curr_id), 
                   post_all[post_all$ID == curr_id, "Students.with.Scores"], "*")
  matches <- c(matches, new_val)
  match_scores <- c(match_scores, scores)
}
all$post_passed <- matches
all$post_numscores <- match_scores

# drop *'s
names(all)[names(all) == "passed"] <- "pre_passed"
all <- all[all$post_passed!="*",]

all <- all[all$post_numscores!="*",]
```

## Big tau
```{r}
all$total.students <- as.numeric(as.character(all$Students.with.Scores)) + as.numeric(as.character(all$post_numscores))
all$post_passed <- as.numeric(as.character(all$post_passed))
all$pre_passed <- as.numeric(as.character(all$pre_passed))
all$diff <- as.numeric(all$post_passed - all$pre_passed)
T_hat <- weighted.mean(all$diff, as.numeric(all$total.students))
print(T_hat)
```

## Fisher randomization test
```{r}
reps <- 10000
ts <- rep(0,reps)
set.seed(1)

for (i in 1:reps) {
  assigns <- sample(c(-1, 1), size = nrow(all), replace = TRUE)
  dup <- all$diff * assigns
  ts[i] <- weighted.mean(dup, as.numeric(all$total.students))
}
p <- length(ts[ts <= T_hat])/reps
print(min(ts))
print(p)
```

## Wilcox Inference
```{r}
w_results = wilcox.test(all$pre_passed, all$post_passed, paired = TRUE, alternative = "greater")

print(w_results)
```




# All student groups diff in diff: change in math score vs change in english score from 2019 to 2021
## Matching again, based on test ID
```{r}
all_t1 <- all[all$Test.ID==1,]
all_t2 <- all[all$Test.ID==2,]
all_t1$ID <- paste(all_t1$School.Code,all_t1$Grade,sep="")
all_t2$ID <- paste(all_t2$School.Code,all_t2$Grade,sep="")

matches_diff <- list()
matches_n <- list()

for (i in 1:(nrow(all_t1))) {
  curr_id <- all_t1[i,"ID"]
  new_diff <- ifelse(any(all_t2$ID == curr_id), all_t2[all_t2$ID == curr_id, "diff"], "*")
  new_n <- ifelse(any(all_t2$ID == curr_id), 
                   all_t2[all_t2$ID == curr_id, "total.students"], "*")
  matches_diff <- c(matches_diff, new_diff)
  matches_n <- c(matches_n, new_n)
}
all_t1$t2_diff <- matches_diff
all_t1$t2_numscores <- matches_n

# drop *'s
names(all_t1)[names(all_t1) == "diff"] <- "t1_diff"
all_t1 <- all_t1[all_t1$t1_diff!="*" & all_t1$t2_diff!="*",]
```


## Big tau
```{r}
all_t1$total.students <- as.numeric(as.character(all_t1$total.students)) + as.numeric(as.character(all_t1$t2_numscores))
all_t1$t1_diff <- as.numeric(as.character(all_t1$t1_diff))
all_t1$t2_diff <- as.numeric(as.character(all_t1$t2_diff))
all_t1$diffindiff <- as.numeric(all_t1$t1_diff - all_t1$t2_diff)
T_hat <- weighted.mean(all_t1$diffindiff, as.numeric(all_t1$total.students))
print(T_hat)
```



## Fisher randomization test
```{r}
reps <- 10000
ts <- rep(0,reps)
set.seed(1)

for (i in 1:reps) {
  assigns <- sample(c(-1, 1), size = nrow(all), replace = TRUE)
  dup <- all$diff * assigns
  ts[i] <- weighted.mean(dup, as.numeric(all$total.students))
}
p <- length(ts[ts >= T_hat])/reps
print(p)
```

## Wilcox Inference
```{r}
w_results = wilcox.test(all$pre_passed, all$post_passed, paired = TRUE, alternative = "two.sided")
print(w_results)
```






# All student groups diff in diff: change from 2017-2019 vs change from 2019-2021
```{r}
prepre <- read.csv("/Users/NatalieHilderbrand/Downloads/sb_ca2017_all_csv_v2/sb_ca2017_all_csv_v2.txt")
prepre <- prepre %>% filter(County.Code != 0 & District.Code != 0 & School.Code != 0 & Grade != 13)
```

## Cleaning
```{r}
prepre <- prepre[, c("Subgroup.ID","Total.Tested.with.Scores", "Total.Tested.At.Entity.Level", "Test.Id", "Grade", "Test.Year", "School.Code", "Percentage.Standard.Met.and.Above")]
colnames(prepre) <- colnames(post)
```

## All student groups
```{r}
prepre_all <- prepre %>% filter(Student.Group.ID == 1)
prepre_all <- prepre_all[!is.na(prepre_all$passed) & prepre_all$passed!="" & prepre_all$passed!="*",]
prepre_all$ID <- paste(prepre_all$School.Code,prepre_all$Grade,prepre_all$Test.ID,sep="")
```

## Matching on all student groups
```{r}
all <- data.frame(post_all)
matches_n_prepre <- list()
matches_n_pre <- list()
matches_scores_prepre <- list()
matches_scores_pre <- list()
for (i in 1:(nrow(all))) {
  curr_id <- all[i,"ID"]

  matches_n_pre <- c(matches_n_pre, ifelse(any(pre_all$ID == curr_id), 
                   pre_all[pre_all$ID == curr_id, "Students.with.Scores"], "*"))
  matches_scores_pre <- c(matches_scores_pre, ifelse(any(pre_all$ID == curr_id), pre_all[pre_all$ID == curr_id, "passed"], "*"))

  matches_n_prepre <- c(matches_n_prepre, ifelse(any(prepre_all$ID == curr_id), 
                   prepre_all[prepre_all$ID == curr_id, "Students.with.Scores"], "*"))
  matches_scores_prepre <- c(matches_scores_prepre, ifelse(any(prepre_all$ID == curr_id), prepre_all[prepre_all$ID == curr_id, "passed"], "*"))
}

all$pre_passed <- matches_scores_pre
all$pre_numscores <- matches_n_pre
all$prepre_passed <- matches_scores_prepre
all$prepre_numscores <- matches_n_prepre

# drop *'s
names(all)[names(all) == "passed"] <- "post_passed"
all <- all[all$pre_passed!="*" & all$prepre_passed!="*" & all$pre_numscores!="*" & all$prepre_numscores!="*",]
```

## Big tau
```{r}
all$total.students <- as.numeric(as.character(all$Students.with.Scores)) + as.numeric(as.character(all$pre_numscores)) + as.numeric(as.character(all$prepre_numscores))
all$post_passed <- as.numeric(as.character(all$post_passed))
all$pre_passed <- as.numeric(as.character(all$pre_passed))
all$prepre_passed <- as.numeric(as.character(all$prepre_passed))
all$diff1 <- as.numeric(all$pre_passed - all$prepre_passed)
all$diff2 <- as.numeric(all$post_passed - all$pre_passed)
all$diffindiff <- as.numeric(all$diff2-all$diff1)
T_hat <- weighted.mean(all$diffindiff, as.numeric(all$total.students))
print(T_hat)
```

```{r}
diffs <- all$post_passed - all$prepre_passed
nums <- as.numeric(as.character(all$Students.with.Scores))+as.numeric(as.character(all$prepre_numscores))
print(weighted.mean(diffs,nums))
reps <- 10000
ts <- rep(0,reps)
set.seed(1)

for (i in 1:reps) {
  assigns <- sample(c(-1, 1), size = nrow(all), replace = TRUE)
  dup <- diffs * assigns
  ts[i] <- weighted.mean(dup, nums)
}
p <- length(ts[ts <= T_hat])/reps
print(p)
```


## Fisher randomization test
```{r}
reps <- 10000
ts <- rep(0,reps)
set.seed(1)

for (i in 1:reps) {
  assigns <- sample(c(-1, 1), size = nrow(all), replace = TRUE)
  dup <- all$diffindiff * assigns
  ts[i] <- weighted.mean(dup, as.numeric(all$total.students))
}
p <- length(ts[ts <= T_hat])/reps
print(min(ts))
print(p)
```

## Wilcox Inference
```{r}
w_results = wilcox.test(all$diff1, all$diff2, paired = TRUE, alternative = "greater")

print(w_results)
```











# Analysis on race and economic status
```{r}
pre_full_info <- pre %>% filter(Student.Group.ID >= 200 & Student.Group.ID <= 227)
post_full_info <- post %>% filter(Student.Group.ID >= 200 & Student.Group.ID <= 227)

# drop null columns
pre_full_info <- pre_full_info[!is.na(pre_full_info$passed) & pre_full_info$passed!="" & pre_full_info$passed!="*",]
post_full_info <- post_full_info[!is.na(post_all$passed) & post_full_info$passed!="" & post_full_info$passed!="*",]

# add ID for matching with school code, grade, test type
pre_full_info$ID <- paste(pre_full_info$School.Code,pre_full_info$Grade,pre_full_info$Test.ID,pre_full_info$Student.Group.ID,sep="")
post_full_info$ID <- paste(post_full_info$School.Code,post_full_info$Grade,post_full_info$Test.ID,post_full_info$Student.Group.ID,sep="")
```

## Matching
```{r}
all <- data.frame(post_full_info)
match_n <- list()
match_scores <- list()
for (i in 1:(nrow(all))) {
  curr_id <- all[i,"ID"]
  new_val <- ifelse(any(pre_full_info$ID == curr_id), pre_full_info[pre_full_info$ID == curr_id, "passed"], "*")
  scores <- ifelse(any(pre_full_info$ID == curr_id), 
                   pre_full_info[pre_full_info$ID == curr_id, "Students.with.Scores"], "*")
  match_n <- c(match_n, scores)
  match_scores <- c(match_scores, new_val)
}
all$pre_n <- match_n
all$pre_passed <- match_scores

# drop *'s
names(all)[names(all) == "passed"] <- "post_passed"
all <- all[all$pre_passed!="*" & all$pre_n!="*",]
```

## One-hot encoding stuff
```{r}
all$diff <- as.numeric(as.character(all$post_passed)) - as.numeric(as.character(all$pre_passed))
all$Test.ID <- all$Test.ID - 1 # 1 if the test is math, 0 if langauge
all$disadv <- ifelse(all$Student.Group.ID <= 207, 1, 0)
all$black <- ifelse(all$Student.Group.ID == 200 | all$Student.Group.ID == 220, 1, 0)
all$hisp <- ifelse(all$Student.Group.ID == 204 | all$Student.Group.ID == 224, 1, 0)
all$asian <- ifelse(all$Student.Group.ID == 202 | all$Student.Group.ID == 222, 1, 0)
all$twoormore <- ifelse(all$Student.Group.ID == 207 | all$Student.Group.ID == 227, 1, 0)
all$native <- ifelse(all$Student.Group.ID == 205 | all$Student.Group.ID == 203 | all$Student.Group.ID == 201 | all$Student.Group.ID == 225 | all$Student.Group.ID == 223 | all$Student.Group.ID == 221, 1, 0)
```

## Regressing
```{r}
lmout <- lm(diff ~ Test.ID + Grade + black + hisp + asian + native, data=all)
print(summary(lmout))
lmout2 <- lm(as.numeric(as.character(pre_passed)) ~ Test.ID + disadv + black + hisp + asian + native + Grade, data=all)
print(summary(lmout2))
lmout3 <- lm(diff ~ Test.ID*disadv + disadv + Test.ID, data=all)
print(summary(lmout3))
```













# Analysis of economic status
```{r}
pre_ec <- pre %>% filter(Student.Group.ID == 31 | Student.Group.ID == 111)
post_ec <- post %>% filter(Student.Group.ID == 31 | Student.Group.ID == 111)

# drop null columns
pre_ec <- pre_ec[!is.na(pre_ec$passed) & pre_ec$passed!="" & pre_ec$passed!="*",]
post_ec <- post_ec[!is.na(post_ec$passed) & post_ec$passed!="" & post_ec$passed!="*",]

# add ID for matching with school code, grade, test type
pre_ec$ID <- paste(pre_ec$School.Code,pre_ec$Grade,pre_ec$Test.ID,pre_ec$Student.Group.ID,sep="")
post_ec$ID <- paste(post_ec$School.Code,post_ec$Grade,post_ec$Test.ID,post_ec$Student.Group.ID,sep="")
```

## Matching
```{r}
all <- data.frame(post_ec)
match_n <- list()
match_scores <- list()
for (i in 1:(nrow(all))) {
  curr_id <- all[i,"ID"]
  new_val <- ifelse(any(pre_ec$ID == curr_id), pre_ec[pre_ec$ID == curr_id, "passed"], "*")
  scores <- ifelse(any(pre_ec$ID == curr_id), 
                   pre_ec[pre_ec$ID == curr_id, "Students.with.Scores"], "*")
  match_n <- c(match_n, scores)
  match_scores <- c(match_scores, new_val)
}
all$pre_n <- match_n
all$pre_passed <- match_scores

# drop *'s
names(all)[names(all) == "passed"] <- "post_passed"
all <- all[all$pre_passed!="*" & all$pre_n!="*",]
```


## Tau hat
```{r}
all$diff <- as.numeric(as.character(all$pre_passed)) - as.numeric(as.character(all$post_passed))
all$disadv <- ifelse(all$Student.Group.ID == 31, 1, 0)
all$total.students <- as.numeric(as.character(all$pre_n)) + as.numeric(as.character(all$Students.with.Scores))
```

## Matching again
```{r}
fin <- all[all$disadv == 1,]
print(nrow(fin))
adv <- all[all$disadv == 0,]
fin$ID <- paste(fin$School.Code,fin$Grade,fin$Test.ID,sep="")
adv$ID <- paste(adv$School.Code,adv$Grade,adv$Test.ID,sep="")

match_n <- list()
match_scores <- list()
for (i in 1:(nrow(fin))) {
  curr_id <- fin[i,"ID"]
  new_val <- ifelse(any(adv$ID == curr_id), adv[adv$ID == curr_id, "diff"], "*")
  scores <- ifelse(any(adv$ID == curr_id), 
                   adv[adv$ID == curr_id, "total.students"], "*")
  match_n <- c(match_n, scores)
  match_scores <- c(match_scores, new_val)
}
fin$adv_n <- match_n
fin$adv_diff <- match_scores

# drop *'s
names(fin)[names(fin) == "diff"] <- "disadv_diff"
fin <- fin[fin$adv_diff!="*" & fin$disadv_diff!="*",]
```


```{r}
fin$diffindiff <- as.numeric(as.character(fin$disadv_diff)) - as.numeric(as.character(fin$adv_diff))
fin$n <- as.numeric(as.character(fin$total.students)) + as.numeric(as.character(fin$adv_n))
T_hat <- weighted.mean(fin$diffindiff, as.numeric(fin$n))
print(T_hat)
```

```{r}
print(mean(as.numeric(as.character(fin$pre_passed))))
print(mean(as.numeric(as.character(fin$post_passed))))
print(mean(as.numeric(as.character(adv$pre_passed))))
print(mean(as.numeric(as.character(adv$post_passed))))
```

## Fisher randomization test
```{r}
reps <- 10000
ts <- rep(0,reps)
set.seed(1)

for (i in 1:reps) {
  assigns <- sample(c(-1, 1), size = nrow(fin), replace = TRUE)
  dup <- fin$diffindiff * assigns
  ts[i] <- weighted.mean(dup, as.numeric(fin$n))
}
p <- length(ts[ts <= T_hat])/reps
print(min(ts))
print(p)
```










# Cohort analysis
```{r}
pre_ca <- pre %>% filter(Student.Group.ID == 1)
post_ca <- post %>% filter(Student.Group.ID == 1)

# drop null columns
pre_ca <- pre_ca[!is.na(pre_ca$passed) & pre_ca$passed!="" & pre_ca$passed!="*",]
post_ca <- post_ca[!is.na(post_ca$passed) & post_ca$passed!="" & post_ca$passed!="*",]

pre_ca$Cohort <- pre_ca$Grade
post_ca$Cohort <- post_ca$Grade-2

# add ID for matching with school code, grade, test type
pre_ca$ID <- paste(pre_ca$School.Code,pre_ca$Cohort,pre_ca$Test.ID,sep="")
post_ca$ID <- paste(post_ca$School.Code,post_ca$Cohort,post_ca$Test.ID,sep="")
```

## Matching
```{r}
all <- data.frame(post_ca)
match_n <- list()
match_scores <- list()
for (i in 1:(nrow(all))) {
  curr_id <- all[i,"ID"]
  new_val <- ifelse(any(pre_ca$ID == curr_id), pre_ca[pre_ca$ID == curr_id, "passed"], "*")
  scores <- ifelse(any(pre_ca$ID == curr_id), 
                   pre_ca[pre_ca$ID == curr_id, "Students.with.Scores"], "*")
  match_n <- c(match_n, scores)
  match_scores <- c(match_scores, new_val)
}
all$pre_n <- match_n
all$pre_passed <- match_scores

# drop *'s
names(all)[names(all) == "passed"] <- "post_passed"
all <- all[all$pre_passed!="*" & all$pre_n!="*",]
```

## Big tau
```{r}
all$post_passed <- as.numeric(as.character(all$post_passed))
all$pre_passed <- as.numeric(as.character(all$pre_passed))
all$diff <- as.numeric(all$post_passed - all$pre_passed)
T_hat <- weighted.mean(all$diff, as.numeric(all$Students.with.Scores))
print(T_hat)
```

## Fisher randomization test
```{r}
reps <- 10000
ts <- rep(0,reps)
set.seed(1)

for (i in 1:reps) {
  assigns <- sample(c(-1, 1), size = nrow(all), replace = TRUE)
  dup <- all$diff * assigns
  ts[i] <- mean(dup)
}
p <- length(ts[ts <= T_hat])/reps
print(min(ts))
print(p)
```

## Wilcox Inference
```{r}
w_results = wilcox.test(all$pre_passed, all$post_passed, paired = TRUE, alternative = "greater")

print(w_results)
```
